Recent demography drives changes in linked selection across the maize genome

### 2015-11-02 manuscript.tex
intro

\significancetext{Both selection and demographic change play important roles in shaping diversity across the genome, but clear empirical examples of the interplay of these two forces are lacking.
Here we document the combined effects of demography and linked selection on genome-wide diversity in domesticated maize and its wild ancestor teosinte.  We estimate that maize underwent a bottleneck of 5\% of the size of the ancestral teosinte population, but that recent expansion has resulted in a maize population perhaps orders of magnitude larger than teosinte.  We show that positive selection on new genic mutations has had relatively little effect on genetic diversity, but that selection against deleterious mutations has dramatically reduced diversity in and immediately around genes in both taxa.  We find that the relative effect of selection depend qualitatively on the age of the polymorphisms evaluated: while older polymorphisms in maize show more limited effects of linked selection, new mutations instead reflect its larger current size and more efficient purifying selection. Our results demonstrate that a complete understanding of genome-wide patterns of diversity will require careful assessment of both demographic history and the effects of linked selection. }

@@

%\jri{we need a starting paragraph on linked selection and demography. check out slotte 2014 briefings functional genomics for a great review of ideas. then we do a paragraph on why domesticates are great for this (modified version of your below)}
%\mbh{Yes, agree here...think the topics addressed in the manuscript need to be set up as generalizable and not just specific to domesticates.  Crops are a great system because we know a fair bit more about demography and the process of selection than in wild species, but they are not entirely unique.}

\dropcap{T}he genetic properties of populations are determined by a constant interplay between genetic drift and selection. Drift is the consequence of a finite population size, such that random sampling of gametes every generation will incrementally lead to changes in genetic variabiligy \cite{dobzhansky1957}. In other words, the demographic history of a population, which includes its population size fluctuations and mating patterns, has a long-lasting impact on the patterns of DNA variability displayed by that population. Conversely, selection systematically alters a population by favoring particular alleles at the expense of others. In contrast to drift, which results from demographic history, selection is the result of the relationship between genotypes, phenoytpes, and ultimately fitness. A classical approach for studying selection is to remove the genome-wide effects of demography and assume that selection will leave locus-specific evidence \cite{Wall2002, akey2009}. Conversely, demography is often researched by excluding potentially selected sites \cite{voight2005, luikart2003, gutenkunst2009}.

However, recent work has shown that demography and selection do not operate independently to determine genetic variability, in large part because linkage allows the effects of selection to be wide-ranging \cite{li2012,slotte2014}. For example, in \emph{Drosophila} the majority of the genome is
impacted by the effects of selection through linkage
\cite{sella2009}. One technique that has been utilized to estimate demographic parameters independently of selective parameters is to limit analyses to only sites far from genes\cite{gazave2014}, but as \emph{Drosophila} demonstrates, it is
difficult to be certain, without species-specific knowledge, that sites even in gene-poor regions of the genome are not influenced by linked selection. Therefore, characterizing selection, demography or both, first requires an understanding of linked selection. It has already been established that patterns of linked selection are heavily influenced by population size ($N_e$) \cite{cutter2013, slotte2014}, but nearly all species including humans \cite{reich1998}, domesticates \cite{hyten2006, bovine2009}, ecological \cite{storfer2014}, and model species \cite{duchen2013} have experienced recent or ancient changes in $N_e$. Although much is known about how the long-term average $N_e$ affects linked selection, much less is understood about the immediate effects and dynamics of changing in $N_e$ on patterns of linked selection.

Because of its relatively simple demographic history and abundant availability of genomic tools, maize (\emph{Zea mays}) represents an excellent organism to study this phenomenon. Maize is a species of tremendous importance worldwide as both a model species \cite{strable2009} and a staple crop \cite{shiferaw2011}. Archaeological and genetic studies have established that maize domestication began in
Central Mexico at least 9,000 years bp
\cite{smith1995,matsuoka2002}, and involved a population bottleneck followed by recent expansion \cite{wright2005}. Because of this simple but dynamic demographic history, patterns of linked selection in wild and domesticated maize can be used to understand the effects of changing $N_e$ on linked selection. Teosinte, which remains extant throughout much of the Americas \cite{wilkes1967}, is the wild ancestor to maize. Additionally, several large-effect
domestication loci \cite{doebley1995, wills2013, wang2015} and putative domestication
regions \cite{hufford2012} have been identified in maize. But despite all that is
known about maize domestication, the demographic parameters of the
domestication process remain uncertain. Specifically, the size of the
maize domestication bottleneck has not been estimated independently of
the bottleneck's duration, nor are there sequence-based estimates of the effective
population size of modern maize. At the same time, the predominant form or forms of selection that have taken place during and since domestication, and how linkage impacts the evidence and patterns of selection across the genome, remain unknown.

For this study, we begin by estimating the parameters of maize domestication based on whole-genome sequencing data. Next, we utilized these estimates to 1) investigate the relative importance of different forms of selection on genetic
variability in the recent and acient past; and 2) research the impact that demographic history, specifically changes in $N_e$, has had on linked selection in maize, and how this compares to the impact of a different demographic history in teosinte. Gaining an improved understanding of the impacts of a dynamic demography on whole-genome variability in maize will help to increase the understanding of these phenomena across all organisms, since changes in $N_e$ are pervasive.

>>>>

\significancetext{Both selection and demographic change play important roles in shaping diversity across the genome, but clear empirical examples of the interplay of these two forces are lacking.
Here we document the combined effects of demography and linked selection on genome-wide diversity in domesticated maize and its wild ancestor teosinte.  We estimate that maize underwent a bottleneck to  $\approx 5\%$ of the size of the ancestral teosinte population, but that recent expansion has resulted in a maize population perhaps orders of magnitude larger than teosinte.  We show that positive selection on new genic mutations has had relatively little effect on genetic diversity, but that selection against deleterious mutations has dramatically reduced diversity in and immediately around genes in both taxa.  We find that the relative effect of selection depends qualitatively on the age of the polymorphisms evaluated: while older polymorphisms in maize show more limited effects of linked selection, new mutations instead reflect its larger current size and more efficient selection. Our results demonstrate that a complete understanding of genome-wide patterns of diversity will require careful assessment of both demographic history and the effects of linked selection. }

@@

\dropcap{T}he genetic diversity of populations is determined by a constant interplay between genetic drift and selection. 
Drift is a consequence of a finite population size and the random sampling of gametes every generation \cite{dobzhansky1957}. 
In contrast to the stochastic effects of drift, selection systematically alters allele frequencies by favoring particular alleles at the expense of others as a result of their effects on fitness.
Researchers often seek to study drift by excluding potentially selected sites \cite{voight2005, luikart2003, gutenkunst2009}, or to study selection by focusing on site-specific patterns under the assumption that genome-wide diversity reflects only the action of drift \cite{Wall2002, akey2009}. 
\jri{careful here with these citations. we're effectively calling these people out, so make sure they're really doing what you say they are!}

Drift and selection do not operate independently to determine genetic variability, however, in large part because linkage allows the effects of selection to be wide-ranging \cite{smith1974, li2012,slotte2014}. 
Linked selection can take the form of hitch-hiking, when the frequency of a neutral allele changes as a result of positive selection at a physically linked site \cite{smith1974}, or background selection, where diversity is reduced at loci linked to a site undergoing selection against deleterious alleles \cite{charlesworth1993}. 
Recent work in a \emph{Drosophila}, for example, has shown that essentially the entire genome is impacted by the combined effects of these processes \cite{sella2009}. \jri{add citations for el yashiv biorxiv and andolfatto nature 2005}
The impact of linked selection, in turn, is heavily influenced by the effective population size ($N_e$), as the  efficiency of natural selection is proportional to the product $N_es$, where $s$ is the strength of selection on a variant \cite{cutter2013, slotte2014}. \jri{add corbett-detig and leffler DOI: 10.1371/journal.pbio.1001388}.
The effective size of a population is not static, and nearly all species, including flies \cite{duchen2013}, humans \cite{reich1998}, domesticates \cite{hyten2006, bovine2009}, and non-model species \citex \jri{this storfer ref didn't seem so great. any review on demography in non-models? or maybe list of several examples?} have experienced recent or ancient changes in $N_e$. 
Although much is known about how the long-term average $N_e$ affects linked selection, much less is understood about the immediate effects and dynamics of changing in $N_e$ on patterns of linked selection.

Because of its relatively simple demographic history and the abundant availability of genomic tools, maize (\emph{Zea mays}) represents an excellent organism to study these effects. 
Maize is a species of tremendous importance worldwide as both a model species \cite{strable2009} and a staple crop \cite{shiferaw2011}. 
Archaeological and genetic studies have established that maize domestication began in
Central Mexico at least 9,000 years bp \cite{smith1995,matsuoka2002}, and involved a population bottleneck followed by recent expansion \cite{wright2005}.\jri{cite eyre-walker 1996 and tenaillon 2004 too}
Because of this simple but dynamic demographic history, patterns of linked selection domesticated maize and its wild ancestor teosinte can be used to understand the effects of changing $N_e$ on linked selection. 
Here, we take advantage of this system to study how recent demographic change impacts patterns of linked selection genome-wide.
We begin by estimating the parameters of the maize domestication bottleneck using whole-genome resequencing data, then investigate the relative importance of different forms of linked selection on diversity in the recent and ancient past. 
We show that while patterns of overall nucleotide diversity reflect long-term differences in $Ne$, recent growth following domestication qualitatively changes these effects, illustrating the importance of understanding demographic history when considering the effects of selection on the genome.
